Demography and selection since maize domestication

### 2015-05-21 manuscript.tex
updates

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious
mutations arising in a population are continuously selected against. When this
form of selection is operating, it can serve to reduce genetic
variability at linked neutral sites, a phenomenon referred called
background selection \cite{charlesworth1993}. Purifying and background selection lead
to lower diversity within genes and other functional sites relative
to neutral regions \textcolor{red}{(ref here)}. We investigated purifying
selection in maize and teosinte by evaluating the average magnitude of reduced
diversity within genes and recovery away from genes in both taxa. When standardized by neutral
levels (which we defined as diversity at least \textcolor{red}{0.01} cM distal
from genes), a stronger reduction of diversity
and slower recovery was observed for teosinte than for maize, implying
that purifying selection has left a more pronounced signature in the
teosinte genome (Figure \ref{purify}). This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication would
have caused enhanced purifying selection for maize. We therefore
conducted the same analysis based on singleton diversity. As a class,
singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. When
evaluating the data in this manner, an opposite relationship was
observed. Maize singleton diversity was lower than teosinte singleton
diversity near genes, and recovered more slowly
(Figure \ref{purify}), implying that in the
recent past maize has been more influenced by purifying selection than
teosinte. Together, these findings imply that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent past. \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure}}
\label{purify}
\end{figure*}

@@

\subsection{Evaluating diversity around substitutions}
To investigate diversity around substitutions, maize and teosinte pairwise diversity was first
calculated in 1,000 kb non-overlapping windows using ANGSD
\cite{korneliussen2014}. This was performed separately for both maize and
teosinte, using the same filters as employed for estimating the
SFS. Next, SNPs and genotypes among maize, teosinte, and \emph{Tripsacum} were called. \emph{Tripsacum}\emph{Tripsacum} bam files
were downloaded from \url{}(TRIPSICUM FILES), and then all SNPs with a
p-value less than 1e-6 were called using ANGSD. Quality filters were
as the same as before, and genotypes were only called when the
posterior probability was above 0.95. From the set of called SNPs and
genotypes, substitutions between maize and tripsicum, as well as
between teosinte and \emph{Tripsacum} were identified using R \cite{R2014} as all positions with
no more than 20\% missing data for which every maize or teosinte
allele differed from the observed \emph{Tripsacum} allele. At each class of
substitution, effects were estimated using the ensembl variant effects
predictor \cite{mclaren2010}.

@@

\subsection{Evaluating diversity around genes and conserved sequences}
Two types of diversity surrounding genes were investigated. The first
was pairwise diversity in 1kb windows, as described previously. The
second was singleton diversity in 1kb windows. Singletons represent
the rarest class of alleles that this dataset can identify, and
collectively demonstrate the most recent patterns of evolution. Minor
allele frequencies were estimated with ANGSD \cite{korneliussen2014} using the
same quality filters previously described. Then, the number of
singletons in each non-overlapping 1kb window was calculated with R
\cite{R2014}. BiomaRt \cite{durinck2009, durinck2005} was then used to identify
the center of each gene. Next, the distance from each diversity window
to the nearest gene center was computed. Teosinte diversity is
generally higher then maize diversity. Therefore, to enable comparisons between
the reduction of diversity around genes in maize and teosinte, a
neutral measure of pairwise and singleton diversity for each taxa was estimated according to
mean pairwise and singleton diversity at windows greater than 0.01 cM from the nearest
gene. Then, pairwise and singleton diversity at each window was
standardized by dividing by the corresponding neutral
measure. Separately for pairwise and singleton diversity in maize and
teosinte, cubic smoothing splines were fit to
describe diversity levels according to the distance to the nearest
gene. Significant differences were assessed by taking 100 bootstrap
samples and re-fitting the cubic smoothing spline to each. Then, the
2.5\% and 97.5\% quantiles of values along the bootstrapped splines
were identified.

>>>>

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious
mutations arising in a population are continuously selected against. When this
form of selection is operating, it can serve to reduce genetic
variability at linked neutral sites, a phenomenon called
background selection \cite{charlesworth1993}. Purifying and background selection lead
to lower diversity within genes and other functional sites relative
to neutral regions \textcolor{red}{(ref here)}. We investigated purifying
selection in maize and teosinte by evaluating the average magnitude of reduced
diversity within genes and recovery away from genes in both taxa. When standardized by neutral
levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity
and slower recovery was observed for teosinte than for maize, implying
that purifying selection has left a more pronounced signature in the
teosinte genome (Figure \ref{purify}). This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may
have elevated the intensity of purifying selection for maize. We therefore
conducted a parallel analysis based on singleton diversity. As a class,
singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Maize was downsampled
to reflect the expected level of singletons from a sample size of 13,
equivalent to that of teosinte, and for singletons our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings imply that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent past. \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure}}
\label{purify}
\end{figure*}

@@

\subsection{Evaluating diversity around substitutions}
To investigate diversity around substitutions, maize and teosinte pairwise diversity was first
calculated in 1,000 kb non-overlapping windows using ANGSD
\cite{korneliussen2014}. This was performed separately for both maize and
teosinte, using the same filters as employed for estimating the
SFS. Next, SNPs and genotypes among maize, teosinte, and \emph{Tripsacum} were called. \emph{Tripsacum} bam files
were downloaded from \url{}(TRIPSICUM FILES), and then all SNPs with a
p-value less than 1e-6 were called using ANGSD. Quality filters were
as the same as before, and genotypes were only called when the
posterior probability was above 0.95. From the set of called SNPs and
genotypes, substitutions between maize and tripsicum, as well as
between teosinte and \emph{Tripsacum} were identified using R \cite{R2014} as all positions with
no more than 20\% missing data for which every maize or teosinte
allele differed from the observed \emph{Tripsacum} allele. At each class of
substitution, effects were estimated using the ensembl variant effects
predictor \cite{mclaren2010}.

@@

\subsection{Evaluating diversity around genes}
Two types of diversity surrounding genes were investigated. The first
was pairwise diversity in 1kb windows, as described previously. The
second was singleton diversity in 1kb windows. Singletons represent
the rarest class of alleles that this dataset can identify, and
collectively demonstrate the most recent patterns of evolution. Minor
allele frequencies were estimated with ANGSD \cite{korneliussen2014} using the
same quality filters previously described. Then, the number of
singletons in each non-overlapping 1kb window was calculated with R
\cite{R2014}. For maize, we generated a parallel set of downsampled
singleton data, for which binomial sampling within R was conducted
based on allele frequencies at each site, to generate a singleton
psuedo dataset of sample size 13, equivalent to our sample size for
teosinte. BiomaRt \cite{durinck2009, durinck2005} was then used to identify
the center of each gene. Next, the distance from each diversity window center
to the nearest gene center was computed.

Teosinte diversity is
generally higher then maize diversity. Therefore, to enable comparisons between
the reduction of diversity around genes in maize and teosinte, a
neutral measure of pairwise and singleton diversity for each taxa was
estimated. For pairwise diversity, this nuetral measure was defined according to
mean pairwise diversity at windows greater than 0.01 cM from the nearest
gene. For singletons, both maize and teosinte still showed reduced
diversity at a distance of 1 cM, so neutral diversity in this case was
defined as mean singleton diversity at positions 0.02 cM or greater
from genes. Then, pairwise and singleton diversity at each window was
standardized by dividing by the corresponding neutral
measure. Separately for pairwise and singleton diversity in maize and
teosinte, cubic smoothing splines were fit to
describe diversity levels according to the distance to the nearest
gene. Significant differences were assessed by taking 100 bootstrap
samples from each set of diversity windows and re-fitting the cubic smoothing spline to each. Then, the
2.5\% and 97.5\% quantiles of values along the bootstrapped splines
were identified.
